A North Sámi to South Sámi machine translation prototype

### 2014-02-08 paper/paper.tex
Added some examples to Fran's part

Norwegian North Germanic. The Saami languages constitute the
westernmost branch of the Uralic language family. They contain most of
the classical Uralic characteristica: A rich verbal morphology (3
numbers and persons, a rich repertoire of infinite forms), and a
medium-size case system with both grammatical and adverbial functions,
and no gender distinction. They also show extensive contact with their
Germanic neighbours. Many grammatical structures ahre head-initial
constructions, as compared to the classical Uralic pattern, and the
tense system is compatible with the North Germanic one.

@@

\item problems caused by this model: ex from two terms to one term to two terms

Example: nob-1,2   sme-x   sma-1,2

Little nob-sma lexical resources, and total lack of sme-sma, we have to find words by comparing translated texts

@@

\item larger syntactic differences between SL an TL than usually done in Apertium?

>>>>

Norwegian North Germanic. The Saami languages constitute the
westernmost branch of the Uralic language family. They contain most of
the classical Uralic characteristica: A rich verbal morphology (3
numbers and persons, a rich repertoire of infinite forms), and a
medium-size case system with both grammatical and adverbial functions,
and no gender distinction. They also show extensive contact with their
Germanic neighbours. Many grammatical structures are head-initial
constructions, as compared to the classical Uralic pattern, and the
tense system is compatible with the North Germanic one.

@@

\subsubsection{Problems caused by this model: ex from two terms to one term to two terms}

Example: nob-1,2   sme-x   sma-1,2
example the verbs ‘read, count, say’, nob: \textit{lese, telle, uttale}, sme: \textit{lohkat} can be used about all three, sma: \textit{lohkedh, ryöknehtidh, jiehtedh}

Little nob-sma lexical resources, and total lack of sme-sma, we have to find words by comparing translated texts (nob-sme with nob-sma)

\subsubsection{Problems combining sma-nob and sme-nob dictionary}
This was done, it had to be manually edited because of nob-words with more than one meaning. 
Example: nob \textit{regne} means both 'to rain' and ‘to calculate’, 
which are two different verbs both in sme \textit{arvit, rehkenastit} and in sma \textit{abrodh, ryöknedh}

@@

larger syntactic differences between SL an TL than usually done in Apertium? \\
This also has consequences for the bidix, because there are many examples of sme-verb has go be translated into object+verb og adverb+verb in sma.
ex: nob ‘å trekke lodd’ sme \textit{vuorbádit} sma \textit{vuerpiem giesedh}, 
nob ‘å presentere’ sme \textit{ovdanbuktit} sma \textit{åvtese buektedh}. 
I mange tilfeller er grunnen til denne  forskjellen at man på sme har fått utviklet flere termer som matcher de norske teremene, mens man på sma
fremdeles må forklare innholdet.
